---
title: "Project 6"
author: "Si-Plus"
date: "2022-11-06"
output: html_document
bibliography: Project 6 Bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- for comments -->
<style>
div.comment pre { background-color:#ffbbbb; }
div.comment pre.r { background-color:#ffbbbb; }
div.comment {color: red;}
</style>

# Introduction

Lepidoptera, the order of butterflies and moths within the class Insecta, comprises an enormous number of species, roughly 160,000 extant species. The unique ability to fly gives this order of species a particularly important role in an ecosystem. To better understand this functionality, the paired wings (i.e., forewings and hindwings) of Lepidoptera are an important area of research. Given the vitality of flight to these species, it is clear that natural selection must play a role in the physical characteristics of wings. Previous work on Lepidoptera wing size and shape has found that neutral selection, sexual selection, and selection due to predator avoidance, as well as physical elements required for flight, may have all played a role in the evolution of fore- and hindwings [@sourakov2013exotic], [@ponce2015molecular], [@rubin2018evolution]. One interesting finding in the literature is that butterflies are still able to fly if their hindwings are removed, but this removal compromises speed and maneuverability of flight [@jantzen2008hindwings]. 

The studies related to experimental manipulations are limited when determining the relationship between shape variables and ecological impact of fore- and hind wings over time. Studies have been conducted to determine whether fore- and hindwing shape and size differs among various butterfly species. Whilst shape of hindwings were found to be more variable than forewings within certain butterfly species [@strauss1990patterns], another experiment demonstrates strong correlation between forewing and hindwing shapes [@chazot2016morpho]. Thus to accurately prove the relationship among fore- and hindwing shapes, and their variance and significance to the butterfly lineage, this can also provide further insight into the evolution of both butterflies and moths, specifically how shape is related to motion, patterns in evolutionary rates, and similarities and differences within varying species. 

The overall goal of this study is to analyze and compare the shapes as well as evolutionary rates of hind- and forewings of Lepidoptera, which will be done through a morphometric approach. The specific questions this study is aiming to answer are:
  1. Are the rates of wing-shape evolution different between the hind- and forewing? By how much?
  2. Have major lineages of the Lepidoptera undergone significant shifts in evolutionary rate?
  3. Are hind- and forewing shapes correlated?


<div class="comment">
Great intro!

</div>

# Methods

## Overview
In order to answer these questions about the wing structure of Lepidoptera and how it has evolved over time, we digitized wing shapes from images of butterfly samples using imageJ. We used the polygon selection tool to trace the forewing and hindwing of each sample after selecting the clearest image with the least damaged wings (up to 3 for each species). To examine shape differences between species we decided to use an outline-based shape analysis, elliptical Fourier analysis (EFA), in which many ellipses are fit to a shape to gain around 99% accuracy; these ellipses can then be described in a dataset. The R package Momocs was used for both this analysis and a Principal Component Analysis (PCA). The PCA will distill many variables attributed to the shapes to fewer variables that describe the most important variance between all the original variables.
 
 
## Analysis Process
Once outlines were created for both forewings and hindwings, a Procrustes transformation was used to scale wings, superimpose them on top of each other, and align them to an optimally determined mean shape. Next, the EFA and PCA were performed on both the hindwings and the forewings. Species information was added to the PCA data and a phylogenetic tree based on those sampled in @kawahara2019phylogen was uploaded for a comparative analysis. In order to calculate the likelihood of rates of morphological evolution across a phylogenetic tree of the Lepidoptera species in our sample, a “non censored test” was performed. This allowed us to find and compare the evolutionary rates of both the hind- and forewings. Next, we used a phylogenetic ridge regression to calculate changes in rates so we could later analyze whether individual clades evolve at different rates compared to the rest of the tree [@kratch2014ridge]. Using search.shift() function, we were able to learn which clades had significantly large rates, where the probability is >0.975 and significantly small rates, where the probability is < 0.025. Finally, we performed a phylogenetic independent contrasts (PIC) analysis to transform mean values for a set of species into values that are statistically independent so that we could regress them against one another. Using a linear model we were able to determine whether or not there was a significant relationship and whether or not hind- and forewing shapes were correlated.

<div class="comment">
Great methods section, too!

</div>


# Results

```{r,include=FALSE}
library(tidyverse)

library(Momocs)

f <- list.files("class_out_data",pattern=".txt|.csv",full.names = TRUE)


out <- read_delim(f[1],delim="\t") %>% 
  as.matrix()

out %>% 
  list() %>% 
  Out() %>% 
  coo_flipx() %>% 
  stack()


out.df <- vroom::vroom(f, id = "filename")

out.df <- out.df %>% 
  mutate(wing=gsub("XY_.+_(hindwing|forewing)\\..+","\\1",basename(filename))) %>% 
  na.omit()

outs.l <- sapply(f,function(x) out.df %>% filter(filename==x) %>% select(X,Y) %>% as.matrix)

wings <- gsub("XY_.+_(hindwing|forewing)\\..+","\\1",basename(names(outs.l)))

outs <-  outs.l %>% 
  Out(fac=list(wing=wings)) %>% 
  coo_flipx()

forewings <- outs %>% 
  filter(wing=="forewing")

hindwings <- outs %>% 
  filter(wing=="hindwing")

forewings %>% 
  stack()

hindwings %>% 
  stack()
```

## Comparative Analysis

```{r,include=FALSE}
fore.min <- forewings %>% 
  coo_nb() %>% 
  min()

forewings %>%
  coo_interpolate(fore.min) %>% 
  fgProcrustes() %>% 
  stack()

hind.min <- hindwings %>% 
  coo_nb() %>% 
  min()

hindwings %>% 
  coo_interpolate(hind.min) %>% 
  coo_slide(id=1) %>% 
  coo_align()  %>%
  fgProcrustes() %>%
  stack()

forewings %>%
  coo_interpolate(fore.min) %>% 
  coo_align()  %>%
  fgProcrustes() %>% 
  efourier(norm=FALSE) 

hindwings %>% 
  coo_interpolate(hind.min) %>% 
  coo_align()  %>%
  fgProcrustes() %>% 
  efourier(norm=FALSE) 

forewing.pca <- forewings %>%
  coo_interpolate(fore.min) %>%
  coo_align()  %>%
  coo_slide(id=1) %>% 
  fgProcrustes() %>% 
  efourier(norm=FALSE) %>% 
  PCA()

hindwing.pca <-hindwings %>% 
  coo_interpolate(hind.min) %>% 
  coo_align()  %>%
  coo_slide(id=1) %>% 
  fgProcrustes() %>% 
  efourier(norm=FALSE) %>% 
  PCA()

hindwings %>% 
  coo_interpolate(hind.min) %>% 
  coo_align()  %>%
  coo_slide(id=1) %>% 
  fgProcrustes() %>% 
  stack


library(ape)

lep.tree <- ape::read.tree("lep_tree2.tre")


lep.tree <- ladderize(lep.tree)
plot(lep.tree,cex=0.1)

lep.tree$tip.label <- gsub("_"," ",lep.tree$tip.label)

basename(names(outs))[1:5]

lep.sp <- read_csv("lep_image_data.csv")

out.data <- tibble(xy.file=basename(names(outs))) %>% 
  mutate(identifier=gsub("XY_|_hindwing|_forewing|.txt","",xy.file)) %>% 
  left_join(lep.sp)

hindwing.pca2 <-  tibble(xy.file=basename(rownames(hindwing.pca$x)),PC1=hindwing.pca$x[,1],PC2=hindwing.pca$x[,2]) %>% 
  left_join(out.data)

forewing.pca2 <-  tibble(xy.file=basename(rownames(forewing.pca$x)),PC1=forewing.pca$x[,1],PC2=forewing.pca$x[,2])%>% 
  left_join(out.data)
```

### PCA Outline Plots

```{r,"pca plots"}
forewing.pca %>% 
  plot_PCA(title = "forewings")

hindwing.pca %>% 
  plot_PCA(title = "hindwings")
```


## Evolutionary Rates

```{r,include=FALSE}
drops <- lep.tree$tip.label[!lep.tree$tip.label%in%unique(out.data$species)]

lep.tree2 <- drop.tip(lep.tree,drops)

hind.pc1 <- hindwing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull

names(hind.pc1) <-  hindwing.pca2%>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(species)

fore.pc1 <- forewing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(PC1)

names(fore.pc1) <-  forewing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(species)


### PC2s
hind.pc2 <- hindwing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(PC2)

names(hind.pc2) <-  hindwing.pca2%>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>%
  summarize(PC2=mean(PC2)) %>% 
  pull(species)

fore.pc2 <- forewing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(PC2)

names(fore.pc2) <-  forewing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(species)

library(phytools)

forePC1.BM<-brownie.lite(lep.tree2,fore.pc1*10)
hindPC1.BM<-brownie.lite(lep.tree2,hind.pc1*10)

forePC2.BM<-brownie.lite(lep.tree2,fore.pc2*10)
hindPC2.BM<-brownie.lite(lep.tree2,hind.pc2*10)
```

### Evolutionary Rates

```{r,"evolutoinary rates"}
# forewings
forePC1.BM$sig2.single
forePC2.BM$sig2.single

# hindwings
hindPC1.BM$sig2.single
hindPC2.BM$sig2.single
```

## Shifts in Evolutionary Rates

```{r,include=FALSE}
library(RRphylo)

# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# 
# BiocManager::install("ggtree")


library(ggtree)
library(wesanderson)

plot_SS <- function(tre=NULL,SS=NULL,tax=NULL){
  
  
  nodes <- as.numeric(rownames(SS$single.clades))
  
  pal <- wes_palette("Zissou1",n=length(nodes))
  sp <- list()
  for(i in nodes){
    sp.i <- extract.clade(tre,i)$tip.label
    
    #print(head(tax))
    sub.names <- lapply(tax,function(x) x[x%in%sp.i]) 
    
    in.clades <- lapply(sub.names,function(x) length(x)>0) 
    all.of.clade <- lapply(sub.names,function(x) all(sapply(sp.i,function(z) z%in%x))) 
    
    high.clade <- names(sub.names)[last(which(all.of.clade==T))]
    all.clades <- names(sub.names)[which(in.clades==T)]
    crown <- ""
    if(high.clade!=last(names(sub.names))) crown <- "crown-"
    
    sub.clades <- NULL
    if(length(grepl("oidea",all.clades))>0) sub.clades <- all.clades[grepl("oidea",all.clades)]
    
    high.clade2 <- paste0(crown,high.clade,": ",paste0(sub.clades,collapse = "+"))
    sp[[paste0(i)]] <- tibble(n=i,species=sp.i,clade=high.clade2)
    
  }
  
  
  d <- do.call(rbind,sp)%>% 
    rename(label=species) 
  
  d2<- d %>% rename(clade_name=clade) 
  
  p <- ggtree(tre)+ scale_y_reverse()
  
  p$data <- p$data %>% left_join(d) %>% left_join(tibble(node=nodes,SS$single.clades) %>% mutate(shift=ifelse(rate.difference>0,"+","-")))
  
  p <-  p+geom_tiplab(aes(col=clade),geom="text",size=1.2)+
    geom_cladelab(data=d2,mapping=aes(node=n,col=clade_name,label=clade_name),offset=1,size=1.5)+
    geom_hilight(data=d2,mapping = aes(node = n,fill=clade_name),alpha = 0.01)+
    scale_fill_manual(values = pal)+
    scale_color_manual(values = pal)+
    theme(legend.position = "none")+geom_nodepoint(mapping=aes(subset = shift =="-"), size=5, shape=25,fill='blue',color='blue',alpha=0.7)+
    geom_nodepoint(mapping=aes(subset = shift =="+"), size=5, shape=24, fill='red',color='red',alpha=0.7)
  p <- p+xlim(NA,6)
  res <- tibble(n=nodes,SS$single.clades) %>% left_join(d %>% select(n,clade) %>% unique)
  
  return(list(plot=p,res=res))
  
}


tax.names <- readRDS("Lep_classification.RDS")

### hind PC1 shifts 
hindPC1.RR <- RRphylo(tree=lep.tree2,y=hind.pc1)
hindPC1.SS<- search.shift(RR=hindPC1.RR,status.type="clade")
hindPC1.res <- plot_SS(lep.tree2,hindPC1.SS,tax = tax.names)

### hind PC2 shifts
hindPC2.RR <- RRphylo(tree=lep.tree2,y=hind.pc2)
hindPC2.SS<- search.shift(RR=hindPC2.RR,status.type="clade")
hindPC2.res <- plot_SS(lep.tree2,hindPC2.SS,tax = tax.names)

### fore PC1 shifts
forePC1.RR <- RRphylo(tree=lep.tree2,y=fore.pc1)
forePC1.SS<- search.shift(RR=forePC1.RR,status.type="clade")
forePC1.res <- plot_SS(lep.tree2,forePC1.SS,tax = tax.names)

### fore PC2 shifts
forePC2.RR <- RRphylo(tree=lep.tree2,y=fore.pc2)
forePC2.SS<- search.shift(RR=forePC2.RR,status.type="clade")
forePC2.res <- plot_SS(lep.tree2,forePC2.SS,tax = tax.names)

```


```{r,"evolutionary shifts plots"}
hindPC1.res$plot
hindPC1.res$res

hindPC2.res$plot
hindPC2.res$res

forePC1.res$plot
forePC1.res$res

forePC2.res$plot
forePC2.res$res
```

## Shape Evolution Correlation 

```{r,include=FALSE}
hindPC1.pic <- pic(hind.pc1,phy = lep.tree2)
forePC1.pic <- pic(fore.pc1,phy = lep.tree2)

PC1.pic <- tibble(
  hind=hindPC1.pic,
  fore=forePC1.pic
)
```

### PC1 regression 

```{r,"PC1 regression"}
PC1.pic %>% 
  ggplot(aes(x=fore,y=hind))+geom_point()+geom_smooth(method="lm")

summary(lm(hind~fore,PC1.pic))
```

```{r,include=FALSE}
hindPC2.pic <- pic(hind.pc2,phy = lep.tree2)
forePC2.pic <- pic(fore.pc2,phy = lep.tree2)

PC2.pic <- tibble(
  hind=hindPC2.pic,
  fore=forePC2.pic
)
```
a significant correlation between PC2 hind- and fore wings
(R-squared = 0.4442)

### PC2 regression

```{r,"PC2 regression"}
PC2.pic %>% 
  ggplot(aes(x=fore,y=hind))+geom_point()+geom_smooth(method="lm")

summary(lm(hind~fore,PC2.pic))
```
no significant correlation between PC2 hind- and fore wings
(R-squared = -0.001955)


<div class="comment">
The implementation of the analysis is wonderful. However, there are a few missing pieces in the results: fig captions and, more importantly, a narrative. It's not enough to produce figures and numbers, we need to express these results in the text.

```{r}
grade <- 75
grade <- grade-7
```

</div>

# Discussion

Through our analyses, we found that fore- and hindwings of Lepidoptera do vary in their evolutionary rates. After analyzing both principle component 1 (PC1) and principle component 2 (PC2) for each wing, the hindwing was found to have a greater evolutionary rate. In looking at the shape morphology, too, there were obvious differences that led us to hypothesize these results. The PC1 rates for the fore- and hindwing were 0.0065 and 0.0101, respectively; the PC2 rates for the fore- and hindwing were 0.0024 and 0.019, respectively. As seen, the hindwing displays faster evolutionary rates. As far as we know, increased genetic variance within a species allows for evolution to occur at a faster rate, as certain traits are more likely to ‘die off’ and traits more fit for survival become ‘established.’  In relation to Lepidopteras, the reasoning for the faster hindwing evolutionary rate may potentially be quite interesting. The action of the forewings essentially drives butterfly flight, while the hindwings serve both as an extended airfoil during gliding flight and to provide additional power during rapid maneuvers in flapping flight. A potential hypothesis, then, is that the forewing shape became a stable trait and, thus, more likely to be conserved through time. The hindwing, on the other hand, could have been a crux of the exponential magnification of butterfly competition.  This has been seen in other species, too. For instance, @bininda2007forelimb found that the forelimb of tetrapods develops earlier than the hindlimb. The conservative nature of this trait means that changes serve as useful phylogenetic ‘markers’ at higher taxonomic levels in tetrapods. 

<div class="comment">
Great point about the hind and forewing being linked mechanically!

</div>


We also found that some major lineages of the Lepidoptera have undergone significant shifts in evolutionary rate. Specifically, Macroheterocera: noctuoidea and crown-Obtectomera: Papilionoidea for hindwings, and  Macroheterocera: Noctuoidea for forewings. In our phylogenetic ridge regression analyses, these species displayed evolutionary rate that were significantly larger than the rest of the tree (at alpha = 0.05), meaning p > 0.975. This could possibly point to some major evolutionary event affecting these species or perhaps a specific environmental niche which induced faster evolution. This finding prompts further exploration in future studies to determine  the cause of these significant shifts in evolutionary rate.

To address our last question, we found that hind- and fore-wing shapes are significantly correlated. In our phylogenetic independent contrasts (PIC) analysis of our PC1, we found a significant relationship (p<0.05) between hing- and forewing shape with R-squared value of 0.4476. We believe that this could be indicative of evolutionary pressures that exist in the environments of the Lepidoptera; despite differences in function between hind- and forewings, both must retain certain aerodynamic characteristics that aid the species with flight in the specific environment in which it lives.

The results ultimately emphasize the significance of determining the existence of a correlation between hind- and fore wing shape and specific variance in evolutionary rates. An additional study in this area examines the structure, properties, functions of forewing-hindwing coupling of honeybees [@ma2019structure]. This study examined the structure of the fore- and hindwings of honeybees, specifically in regards to mechanism and differences of angle variations. This study demonstrated the strong differences that occur that allow for angle ranges, drag augment, and stroke movement that were shown to develop over time within the honeybee species. One can extrapolated that similar wing-shape advantages may exist for the Lepidoptera, thus leading to correlation of hind- and forewing shape. Moreover, the varying rates of hind- and forewing evolution support the hypothesis that this paired-wing structure allows for differing functions between the hind- and forewings. This exploration of hind- and forewings of the Lepidoptera is, as seen here, an important emerging research niche, as these evolutionary patterns can reveal insights on the functionality and importance of wing structures.

<div class="comment">
Such a great discussion!

</div>


# Author Contributions
Teresa worked on the introduction, bibliography, discussion, and R markdwon.
Brendan worked on the results, research, and discussion. 
Kayla worked on the results, R markdown, and methods.
Courtney worked on the introduction, research, and discussion. 

<div class="comment">
Excellent work! Only missing some narrative to the results. Thanks for staying the course and implementing what is more reflective of graduate-level work. Congrats to you all!

```{r}
print(grade)
```
</div>

# References
