---
title: "Project 7"
author: "Teresa Capella, Courtney Cheung, Brendan Fox, Kayla Vidal"
date: "2022-11-27"
output: html_document
bibliography: Project 7 Bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, packages, include=FALSE}
library(tidyverse)
library(knitr)
library(MuMIn)
```
# Introduction
This project will focus on examining the force-length (FL) relationship involved in muscle contraction. The theoretical background of the FL relationship is based upon the sliding filament theory and the cross-bridge theory. According to the sliding filament theory, length change in sarcomere fibers (and, thus, in muscles) is due to the sliding of actin and myosin filaments between each other, which acts to shorten the overall muscle length. The cross-bridge theory, then, posits that this sliding is made possible by the binding of cross bridges between actin and myosin [@RassierLengthDep]. Together, these theories imply that the generation of force is influenced by the cross bridges made where actin and myosin overlap, and therefore that muscle length can limit the amount of force produced. Previous work suggests that the maximum force will be generated when myosin-actin filament overlap is optimal, and that produced force will decrease as this overlap is further from its optimal state [@RassierLengthDep]. However, [@RassierLengthDep] also discusses the moment arm of the muscle at the target joint as well as the excursion of the muscle as other influencing factors in the production of force in addition to the cross-bridge theory and sliding filament theory.

This analysis will specifically focus on the isometric force produced by the human upper limb at various angles. The questions that we will aim to answer are: (1) Do class-wide force-angle data for isometric MVC accord to a typical FL relationship both under control and fatigued conditions? (2) Is there a significant shift between θmax between the control and fatigue FL relationships? By analyzing class-wide force-angle data, this will allow for comparison of the different angles at which maximum isometric force occurs in different fatigue intensities. This permits for further discussion on the underlying mechanisms that result at times of fatigue. [@EnokaMusFatigue]

# Methods
We answered these questions by collecting various data that measured the isometric force over the range of motion of the forearm produced by maximum voluntary contractions (MVCs). Recording different MVC forces that involve a different elbow angle for thirty seconds allowed for over twelve arm angles, 45-157.5 at 11.25 intervals. To test MVCs after eccentric fatigue, we used a backpack, performing isokinetic controlled drops to fatigue flexors before repeating the previous procedure. Maximum force for each recording and across all angles of both control and fatigued experiments were recorded to determine if any shifts occurred at angle F max and at which angles created the most force. 

```{r, Setup, include = FALSE}
dat.f <- list.files(path="./Project_8_data")
dat.l <- list()

#Setup- Running for loop that hits each data file and compiles the metadata
#Pulls max forces from trial and creates mean
for(i in dat.f){
  metadata <- unlist(strsplit(i,'_'))
  subj <- metadata[2]
  ang <- metadata[3]
  exp <- gsub('.csv','',metadata[4])
  trial.force <- read_csv(paste0('Project_8_data/',i),col_names=FALSE)
  trial.force$X1 <- gsub("[a-zA-Z :]","",trial.force$X1)
  maxforce <- trial.force %>% 
    pull(X1) %>% 
    as.numeric() %>% 
    mean()
  dat.l[[i]] <- tibble(
    subject=subj,
    ang=as.numeric(ang),
    exp=exp,
    force=maxforce
  )
}


dat <- do.call(rbind,dat.l)
dat.p <- dat %>% 
  group_by(exp,subject) %>% 
  mutate(force=force/max(force)) %>%
  print()
  
```

```{r, Model Fitting with AIC Scores, include=FALSE}


dat.p2 <- dat.p %>% 
  na.omit()

AICs <- dat.p2%>%
  group_by(subject,exp)%>%
  summarize(
    m2=AICc(lm(force~poly(ang,2))), #second order
    m3=AICc(lm(force~poly(ang,3))), #third order
    m4=AICc(lm(force~poly(ang,4))) #fourth order
  )%>%
  pivot_longer(m2:m4,names_to="model",values_to="AICc")

x.pred <- seq(45,157.5,length.out = 1000)
fits <- dat.p2%>%
  group_by(subject,exp)%>%
  summarize(
    m2=predict(lm(force~poly(ang,2)),newdata=data.frame(ang=x.pred)), #second order
    m3=predict(lm(force~poly(ang,3)),newdata=data.frame(ang=x.pred)), #third order
    m4=predict(lm(force~poly(ang,4)),newdata=data.frame(ang=x.pred)) #fourth order
  )%>%
  pivot_longer(m2:m4,names_to="model")%>%
  group_by(subject,exp,model)%>%
  summarize(theta_max=x.pred[which.max(value)])

dat.p2 %>%
  ggplot(aes(ang,force,col=exp))+geom_point()


```
# Results

```{r, Visualizing datad, include=TRUE}

dat.p2 %>%
  ggplot(aes(ang,force,col=exp))+geom_point()
```


```{r,Comparing Models, include=FALSE}

best.models <- fits%>%
  left_join(AICs)%>%
  group_by(subject,exp)%>%
  mutate(best=AICc==min(AICc))%>%
  filter(best==TRUE)%>%
  dplyr::select(-best)%>%
  print()
```

```{r,Results and Shift, include=TRUE}
best.models%>%
  pivot_wider(id_cols=subject,names_from = exp,values_from=theta_max)%>%
  mutate(shift=fatigue-control) %>% 
  ungroup() %>%
  na.omit() %>% 
  summarise(mean.shift=mean(shift),se.shift=sd(shift)/sqrt(length(shift)))


```
# Discussion

For both control and fatigued conditions, the class-wide force-angle data revealed typical FL relationships. As the sliding filament theory and crossbridge theory would predict, we observed a 𝜃max (i.e., the angle at which maximum force was achieved) at an intermediate angle, or an intermediate sarcomere length. The crossbridge theory would suggest that this 𝜃max occurs at the optimal myosin-actin overlap, allowing for the maximum number of crossbridges to form and thus generate the most force [@RassierLengthDep]. Additionally, our data supports previous work on the effects of fatigue on the force-length relationship. We observed a mean shift in 𝜃max of 10.9° from control to fatigue conditions; a similar rightward shift post-fatigue of FL behavior has been found in previous studies [@BiewenerFatigue]. 
There was a significant shift, on average +10.9º, in the value of the angle in which the maximum force was achieved (𝜃max). For both control and fatigued experiments, the plateau region, or the area in which crossbridge attachments can optimally form, occurred at a larger angle. However, this region shifted to the right for fatigued experiments. Overall, we believe this indicates that the fatigued muscle has a decrease in tetanic (or sustained) force at shortened muscle lengths. While some other studies have come to a similar conclusion, such as [@GauthierTetanic], other studies including  [@BiewenerFatigue] remain unconvinced and believe that fatigue does not seem to directly result in changes in a muscle’s FL properties at all. More studies should be done to investigate the reasoning behind this relationship both in vivo and in vitro. Many cellular mechanisms of fatigue have been investigated in muscle fibers but there are still varying conclusions as to why these FL relationships occur.  [@TomalkaForceLen]


# Author Contribution
Brendan worked on the code and the results. 
Kayla worked on the code, results, and bibliography. 
Teresa worked on the introduction and discussion. 
Courtney worked on the introduction, methods, and discussion. 

# References